`SCPCL000824` - Using `SingleR` for classification of tumor cells

### 2024-07-05 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
add tumor only and comparison of reference labels

This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
We used three different references with `SingleR`:

1. `Blueprint`: Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
Ambiguous cells are removed from `SCPCL000822`.
2. `Original`: `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
3. `Original and Blueprint`: `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.

@@

```{r, warning=FALSE}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
        x |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

@@

```{r}
# list of columns with singler results for UMAPs
singler_annotation_columns <- c(
  "blueprint_tumor_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Blueprint", "Original", "Original + Blueprint")

singler_annotation_columns |> 
  purrr::imap(\(annotation_column, title){
    ggplot(
    classification_df,
    aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        classification_df, -.data[[annotation_column]]
      ),
      color = "gray80",
      alpha = 0.5,
      size = 0.5
    ) +
    # set points for desired cell type
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(
      vars(.data[[annotation_column]]),
      ncol = 3
    ) +
    scale_color_brewer(palette = "Dark2") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    guides(
      color = guide_legend(
        title = title,
        # more visible points in legend
        override.aes = list(
          alpha = 1,
          size = 1.5
        )
      )
    )
  })
```

@@

First we will just look at the overlap in the tumor cells between all three methods.

```{r}
tumor_cells <- list(
  "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
  "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
  "Original and Blueprint" = classification_df$barcodes[classification_df$updated_singler_blueprint_classification == "Tumor"]
)

@@

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(singler_consensus_call = dplyr::case_when(
    blueprint_tumor_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
    blueprint_tumor_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
    .default = "Normal"
  ))

@@

It looks like the cells that are normal are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 
This is what I would expect given the marker gene expression in these cells.

@@

```{r, fig.height=7}
methods <- c(
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Original", "Original and Blueprint")

# compare blueprint to original and origintal + blueprint
plot_jaccard(classification_df,
             annotation_column = "blueprint_tumor_classification_lumped",
             methods_to_compare = methods,
             column_title = "Blueprint",
             legend_match = "Original") |> 
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )

# compare original to original + blueprint
plot_jaccard(classification_df,
             annotation_column = "updated_singler_classification_lumped",
             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
             column_title = "Original",
             legend_match = "Original")

@@

- There seems to be mostly consensus across all three methods on which cells are tumor cells, which is consistent with the upset plot. 
- When only using Blueprint, a lot of cells are categorized as "Normal". 
This means they match the non-tumor cells identified in `SCPCL000822` more than they match the normal cell types in the `Blueprint` reference. 
Additionally, with only Blueprint, there is a group of cells that are called as tumor but actually appear to match endothelial cells present in `SCPCL000822`.
- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to `erythrocyte` in the reference that just uses `SCPCL000822` (original). 
These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
- When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chrondrocytes. 
We also see that for the most part the top cell types identified are the same except when using blueprint you see the addition of epithelial cells.

@@

I tend to favor using one of these options:

- `Blueprint`: `SCPCL000822` with cells labeled as either tumor or normal in combination with Blueprint. 
- `Original and Blueprint`: `SCPCL000822` with tumor cells labeled and all other cells labeled with the original `SingleR` annotations in combination with `Blueprint`.

I don't like using just the `Original`, where the only reference comes from a single sample, `SCPCL000822`. 
It's very likely that the types of normal cells will differ across samples, so I think we want to be able to use `Blueprint` to help classify any of those cells.

Another option would be to try running with just the tumor cells from `SCPCL000822` as a reference in combination with `Blueprint`, removing the "normal" cells from the reference. 
But either of the above options I think would be sufficient to labeling and pulling out tumor cells. 
Let's confirm this by comparing to our previously used marker gene based methods for classifying this sample.

@@

It looks like using Blueprint alone is the most similar to all marker based methods. 
However, it looks like `UCell` categorizes more cells as normal that are classified as tumor with any of the `SingleR` methods.

For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor. 
This tells me that using `SingleR` may be slightly more conservative than the marker gene based methods.

@@

```{r}
# create a density plot showing the distribution of marker gene expression across classification methods 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker genes from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original",
      method == "updated_singler_blueprint_classification" ~ "Original + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
  )

@@

```{r fig.height=10}
# plot gene set scores for each cell  
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  ) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker genes from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original",
      method == "updated_singler_blueprint_classification" ~ "Original + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
    )

@@

- For the most part, all `SingleR` references identify a similar set of tumor cells. 
However there are some cells that are only identified in one or two methods making the full set of tumor cells unique across references. 
- All three references identify the same top normal cell types with some overlapping sets. 
- There does not appear to be a consensus between the original `SingleR` annotations and the new annotations for any of the references. 
- Using `SCPCL000822` with tumor and normal labels in combination with `Blueprint` seems to be the most consistent with using the marker gene based methods for calling tumor cells. 
Additionally, this reference showed the greatest distinction between tumor cells and all other cells in marker gene expression and gene set scores.

>>>>

This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
We used four different references with `SingleR`:

1. `Blueprint`: Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
Ambiguous cells are removed from `SCPCL000822`.
2. `Tumor only`: Both the full `BlueprintEncodeData` reference from `celldex` and only the tumor cells from `SCPCL000822`. 
3. `Original`: `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
4. `Original and Blueprint`: `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference.

@@

```{r, warning=FALSE}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
          # give NA's a unique label so we can easily track and plot them later
          # they will get grouped with all remaining cell types unless they are a top cell type
        dplyr::if_else(is.na(x), "Not labeled", x) |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

@@

```{r}
# list of columns with singler results for UMAPs
singler_annotation_columns <- c(
  "blueprint_tumor_classification_lumped",
  "tumor_only_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Blueprint", "Tumor only", "Original", "Original + Blueprint")

singler_annotation_columns |> 
  purrr::imap(\(annotation_column, title){
    ggplot(
      classification_df,
      aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
    ) +
      # set points for all "other" points
      geom_point(
        data = dplyr::select(
          classification_df, -.data[[annotation_column]]
        ),
        color = "gray80",
        alpha = 0.5,
        size = 0.1
      ) +
      # set points for desired cell type
      geom_point(size = 0.1, alpha = 0.5) +
      facet_wrap(
        vars(.data[[annotation_column]]),
        ncol = 3
      ) +
      scale_color_brewer(palette = "Dark2") +
      # remove axis numbers and background grid
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL) +
      guides(
        color = guide_legend(
          title = title,
          # more visible points in legend
          override.aes = list(
            alpha = 1,
            size = 1.5
          )
        )
      )
  })
```

@@

First we will just look at the overlap in the tumor cells between all four methods.

```{r}
tumor_cells <- list(
  "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
  "Tumor only" = classification_df$barcodes[classification_df$tumor_only_classification == "Tumor"],
  "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
  "Original and Blueprint" = classification_df$barcodes[classification_df$updated_singler_blueprint_classification == "Tumor"]
)

@@

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(singler_consensus_call = dplyr::case_when(
    blueprint_tumor_classification == "Tumor" & tumor_only_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
    blueprint_tumor_classification == "Tumor" | tumor_only_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
    .default = "Normal"
  ))

@@

It looks like the cells that are normal or ambiguous are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 
This is what I would expect given the marker gene expression in these cells. 
This plot looked different when we didn't include the classifications from the tumor only reference, where most of the smaller clusters were considered normal and not ambiguous. 
It looks like using just the tumor cells from `SCPCL000822` with Blueprint is calling more cells as tumor than the other methods.

@@

```{r, fig.height=7}
methods <- c(
  "tumor_only_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Tumor only", "Original", "Original and Blueprint")

# compare blueprint to tumor, original and original + blueprint
plot_jaccard(classification_df,
             annotation_column = "blueprint_tumor_classification_lumped",
             methods_to_compare = methods,
             column_title = "Blueprint",
             legend_match = "Original") |> 
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )

# compare original to tumor, original + blueprint
subset_methods <- c(
  "tumor_only_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Tumor only", "Original and Blueprint")

plot_jaccard(classification_df,
             annotation_column = "updated_singler_classification_lumped",
             methods_to_compare = subset_methods,
             column_title = "Original",
             legend_match = "Original")

# compare tumor to original and blueprint
plot_jaccard(classification_df,
             annotation_column = "tumor_only_classification_lumped",
             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
             column_title = "Tumor only",
             legend_match = "Tumor only")

@@

- There seems to be mostly consensus across all four methods on which cells are tumor cells, which is consistent with the upset plot. 
- When only using Blueprint, a lot of cells are categorized as "Normal". 
This means they match the non-tumor cells identified in `SCPCL000822` more than they match the normal cell types in the `Blueprint` reference. 
Additionally, with only Blueprint, there is a group of cells that are called as tumor but actually appear to match endothelial cells present in `SCPCL000822`.
- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to `erythrocyte` in the reference that just uses `SCPCL000822` (original). 
These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
- When comparing using just `SCPCL000822` (original) to `SCPCL000822` and Blueprint (original and blueprint), we see an overlap in endothelial cells and chrondrocytes. 
We also see that for the most part the top cell types identified are the same except when using blueprint you see the addition of epithelial cells. 
- Using tumor only appears to label some cells as tumor that are labeled as normal with other methods. 
For example, comparing tumor only to both original and original and blueprint shows that tumor cells are classified as endothelial cells using either of the other two methods.

@@

I tend to favor using the original and Blueprint option that uses tumor cell labels and `SingleR` labels from `SCPCL000822` along with `BlueprintEncodeData`.

It's possible that the Blueprint with tumor and normal cells from `SCPCL000822` is absorbing bad calls into the broader `Normal` category, so we probably want to avoid that. 
The tumor only appears to be calling more cells as tumor than may be we want to be? 
These same cells are categorized as endothelial cells with both Original and Original and blueprint. 
I don't like using just the `Original`, where the only reference comes from a single sample, `SCPCL000822`. 
It's very likely that the types of normal cells will differ across samples, so I think we want to be able to use `Blueprint` to help classify any of those cells.

Let's look into the original and blueprint more, specifically let's see if the non tumor cells tend to align with the normal cells from `SCPCL000822` or the normal cells from `BlueprintEncodeData` more.

```{r}
# label cells with original or blueprint label
classification_df <- classification_df |> 
  dplyr::mutate(
    orig_ref = dplyr::if_else(
      stringr::str_detect(updated_singler_blueprint_ontology, "CL:"),
      "Blueprint", 
      "SCPCL000822"
    ), 
    combined_annotation = glue::glue("{updated_singler_blueprint_classification}-{orig_ref}")
  )
```

Below is a table that shows the number of cells for each cell type that correspond to each reference. 
The table is sorted by the number of cells assigned to the original reference from `SCPCL000822`.

```{r}
# make a table showing which cells are assigned to which reference 
celltype_table <- table(classification_df$updated_singler_blueprint_classification, 
      classification_df$orig_ref)
# sort by SCPCL000822
celltype_table[order(celltype_table[,2], decreasing = TRUE), ]
```

From this table, we see a number of cell normal cell types that have more cells that match the `SCPCL000822` reference over the Blueprint reference including endothelial cells and chondrocytes. 
It looks like fibroblasts and neurons better resemble the cells in Blueprint over `SCPCL000822`. 
And finally, there are a number of cells that are only from Blueprint, like astrocytes and epithelial cells.

Based on the data shown so far, I think using the Original with blueprint would be sufficient to labeling and pulling out tumor cells. 
Let's confirm this by comparing to our previously used marker gene based methods for classifying this sample.

@@

It looks like using Blueprint alone is the most similar to all marker based methods. 
However, it looks like `UCell` categorizes more cells as normal that are classified as tumor with any of the `SingleR` methods. 
Also, using the tumor cells only categorizes more cells as tumor that are classified as normal with other marker gene based methods.

For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor. 
This tells me that using `SingleR` may be slightly more conservative than the marker gene based methods other than when using tumor cells alone.

@@

```{r}
# create a density plot showing the distribution of marker gene expression across classification methods 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker genes from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "tumor_only_classification" ~ "Tumor only",
      method == "updated_singler_classification" ~ "Original",
      method == "updated_singler_blueprint_classification" ~ "Original + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
  )

@@

```{r fig.height=10}
# plot gene set scores for each cell  
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  ) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker genes from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "tumor_only_classification" ~ "Tumor only",
      method == "updated_singler_classification" ~ "Original",
      method == "updated_singler_blueprint_classification" ~ "Original + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
    )

@@

- For the most part, all `SingleR` references identify a similar set of tumor cells. 
However there are some cells that are only identified in one or two methods making the full set of tumor cells unique across references. 
- All four references identify the same top normal cell types with some overlapping sets. 
- There does not appear to be a consensus between the original `SingleR` annotations and the new annotations for any of the references. 
- Using `SCPCL000822` with tumor and normal labels in combination with `Blueprint` seems to be the most consistent with using the marker gene based methods for calling tumor cells. 
Additionally, this reference showed the greatest distinction between tumor cells and all other cells in marker gene expression and gene set scores.
